\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-09-25 analysis/manuscript.tex
Minor language fixes

The controversies surrounding hybrid open access and transformative
agreements have led to varying policy conclusions. For instance, the
UK's Joint Information Systems Committee (JISC) evaluated its open
access strategy, which included transformative agreements (Brayman et
al., 2024). The evaluation revealed that while these agreements had a
significant impact on the growth of open access in the UK, they had
limited effects in facilitating a global shift towards full open access.
As a result, the report advised that the UK's open access strategy
should be reassessed. After making similar observation, Norwegian and
Swedish universities and their consortia also argued for policy changes
(Holden et al., 2023; Widding, 2024). Notably, the Association of
Swedish Higher Education Institutions (Sveriges universitets- och
högskoleförbund, SUHF) recommended only supporting agreements for
publishing in fully open access journals by 2026 (Widding, 2024).
Furthermore, cOAlition S will conclude its financial support of
transformative agreements at the end of 2024, but will continue to view
open access resulting from such agreements as compliant. \footnote{\url{https://www.coalition-s.org/coalition-s-confirms-the-end-of-its-financial-support-for-open-access-publishing-under-transformative-arrangements-after-2024/}}.
cOAlition S also removed the majority of hybrid journals from its
Transformative Journal program in 2023 due to publishers' failure to
meet self-defined open access growth targets (Brainard, 2023). By
contrast, the German DEAL consortium announced a five-year
transformative agreement with Elsevier starting in 2024 and also renewed
its contracts with Springer Nature and Wiley until the end of 2028.
Similarly, the Colombia Consortium signed the first transformative
agreements in Latin America (Muñoz-Vélez et al., 2024).

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the uptake of open access in hybrid
journals, and the extent to which this can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Brayman et al., 2024; Haucap et al., 2021; Huang et al.,
2020; Pölönen et al., 2020; Taubert et al., 2023; Waltman \& Lamers,
2022; Wenaas, 2022) or publisher portfolios (Bakker et al., 2024; Fraser
et al., 2023; Jahn et al., 2022; Momeni et al., 2023; Pieper \&
Broschinski, 2018; Schmal, 2024), while large-scale studies relied on
self-reported agreement data (Moskovkin et al., 2022), or used APC
pricing lists (Shu \& Larivière, 2023). In particular, data availability
is a limiting factor when studying the impact of transformative
agreements (Bakker et al., 2024), because bibliometric databases, even
though many allow the retrieval of open access articles in hybrid
journals, do not directly attribute them to specific transformative
agreements. Likewise, article-level open access invoicing and cost data
are only partly available, which would make it possible to establish a
direct link between transformative agreements and open access
publications (Jahn et al., 2022; Kramer, 2024).

@@

This study combines data from multiple publicly available sources, as
shown in Figure \ref{fig:data_workflow}. Initially, transformative
agreement data retrieved from cOAlition S Journal Checker
Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
provided information about journal portfolios and participating
institutions. After identifying hybrid journals by excluding fully open
access journals, Crossref (Hendricks et al., 2020) served as the primary
data source for article-level metadata including Creative Commons (CC)
license information to indicate open access availability on publisher
websites. Because of a lack of comprehensive publicly available
invoicing data, open access articles published through transformative
agreements were determined by first author affiliations from OpenAlex
(Priem et al., 2022), linked to eligible institutions according to the
transformative agreement data. In the following, the steps are described
in detail.

@@

Data gathering started with obtaining journals included in
transformative agreements from the publicly available Transformative
Agreement Data dump\footnote{\url{https://journalcheckertool.org/transformative-agreements/}}
used by the cOAlition S Journal Checker Tool \footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}},
a voluntary effort based on publicly disclosed contracts in the ESAC
Transformative Agreement Registry.\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}.
The dump consists of multiple online Google spreadsheets where each data
file represents one agreement listed in the ESAC Transformative
Agreement Registry. From the retrieved spreadsheet files, journals and
institutions involved per agreement were obtained.

It should be noted that although many library consortia see the need to
register their agreements through the ESAC registry, many fail to
disclose all details, including contracts (Bakker et al., 2024; Kramer,
2024). Furthermore, the extent to which ESAC is comprehensive remains
uncertain, potentially limiting the coverage of transformative
agreements in the Journal Checker Tool. Another limitation of using the
Journal Checker Tool and its underlying publicly available data dump to
study the development of transformative agreements over time is that
expired transformative agreements are constantly removed. To address
this, four different snapshots were safeguarded and combined for this
study: self-archived versions from July 2021, July 2022, and May 2023,
as well as the most current dump downloaded on 11 December 2023. This
ensured that transformative agreements, which ended from 2021 onwards,
were included, representing the majority of transformative agreements.
Overall, the four combined Transformative Agreement Data dumps used in
this study contained 729 of 869 agreements listed in the ESAC
Transformative Agreement Registry by December 2023.

@@

Affiliation metadata about corresponding authors are crucial for the
planning and evaluation of transformative agreements because they are
considered responsible for arranging open access publication (Borrego et
al., 2021; Geschuhn \& Stone, 2017; Schimmer et al., 2015). For this
study, country and institutional affiliations were retrieved from
OpenAlex. Because the corresponding authorship field was not fully
supported by OpenAlex at the time of analysis, and respective
affiliation data were only available for 54\% of the investigated
articles, this study focused on first authors and their affiliations
instead; approximately 90\% of the articles examined had first author
affiliation metadata in OpenAlex, which is a much larger proportion than
previously reported for the October 2022 snapshot (Zhang et al., 2024).
First authors typically contribute the most to a paper and are often
considered lead authors (Larivière et al., 2016), although notable
differences in corresponding author metadata were observed across
subjects, publication years and bibliometric databases
(Chinchilla-Rodríguez et al., 2024). In case of lacking data about
corresponding authors in bibliometric databases, related studies also
utilised first authors as a proxy to examine open access payments and
transformative agreements (Haucap et al., 2021; Shu \& Larivière, 2023;
Zhang et al., 2022).

To estimate the impact of transformative agreements on hybrid open
access, participating institutions from the Transformative Agreement
Data dump, which the cOAlition S crowd-sourced from agreement documents
and consortia, were matched with the first author affiliations recorded
by OpenAlex using the ROR-ID. Matching also considered the duration of
agreements according to the ESAC registry. 502 agreements were active
between 2018 and 2022. Upon inspection, Transformative Agreement Data
did not comprehensively cover associated institutions, such as
university hospitals or institutes of large research organisations such
as the Max Planck Society. To improve the matching, Transformative
Agreement Data were automatically enriched with ROR-IDs from associated
organisations according to OpenAlex's institution entity data.

@@

Between 2018 and 2022, 11,189 out of 12,857 eligible hybrid journals in
transformative agreements published at least one open access article
under a Creative Commons license. These 11,189 hybrid journals
constituted the foundation for the subsequent analyses. Collectively,
they provided open access to 742,369 out of 8,146,958 articles during
the investigated period, representing a five-year open access proportion
of 9.1\%. Authors who could make use of transformative agreements at the
time of publication contributed 328,957 (44\%) open access articles.
Overall, this investigation was able to establish a link between open
access articles and eligible institutions for 394 out of 502 (78\%)
active transformative agreements between 2018 and 2022 through the
analysis of publication metadata, including affiliations and agreement
data.

@@

Figure \ref{fig:results_overview}C depicts the substantial variations
among the hybrid journals included in transformative agreements in terms
of open access uptake. Although the median generally follows the trend
shown in Figure \ref{fig:results_overview}A, the farther stretch of
upper quartiles and whiskers over the years illustrates that an
increasing number of journals published an above-average proportion of
open access articles. In 2022, 25\% of hybrid journals (n = 2,576) had
an open access uptake of 29\%, and 6.6\% of journals (n = 744) provided
the majority of their articles under a Creative Commons license in the
same year. On average, these journals were smaller (M = 75, SD = 186)
than those with an open access share below 50\% (M = 164, SD = 347).

@@

However, there are notable differences between the three large
publishers. Although Elsevier published the largest volume of articles
(n = 2,770,826, 34\%), it recorded a comparably low number of open
access articles, including those that are associated with transformative
agreements. In contrast, Springer Nature and Wiley provided open access
to a larger proportion of their articles (13\% of Springer Nature
articles and 15\% of Wiley articles were open access), leading to higher
open access market share (23\% Springer Nature resp. 23\% Wiley). This
difference between Elsevier on the one hand and Springer Nature and
Wiley on the other can be attributed to transformative agreements, as
the latter made the majority of their open access articles available
through such deals (57\% Springer Nature resp. 55\% Wiley).

@@

The varying degrees of adoption of open access across the three major
publishers can be attributed to distinct approaches to transformative
agreements. Springer Nature, for example, started offering open access
agreements for its hybrid journal portfolio to selected consortia such
as the Max Planck Society, the Swedish Bibsam Consortium and the Finnish
FinELib consortium in 2015 under the name Springer Compact\footnote{\url{https://web.archive.org/web/20180414062853id_/http://www.liber2015.org.uk/wp-content/uploads/2015/03/Springer-Compact.pdf}}.
Prior to this, Springer had some pilot agreements with a small number of
institutions, including the University of Göttingen (Schmidt \& Shearer,
2012). However, the Springer Compact agreements were not included in the
data as they concluded before the start of the transformative agreement
data collection in June 2021. Nonetheless, the results demonstrate the
importance of agreements for Springer Nature's hybrid open access
business over the past five years (Figure \ref{fig:publisher_figure}).
In 2022, 66\% (n = 34,725) of open access in Springer Nature hybrid
journals was enabled through transformative agreements. In the same
year, 70\% (n = 37,316) of Wiley's open access articles could be linked
to transformative agreements. By contrast, Elsevier published a
comparatively lower proportion of its open access articles through
transformative agreements in 2022 (n = 32,627; 54\%).

@@

Figure \ref{fig:subject_panel} presents the relative growth of hybrid
open access by subject area between 2018 and 2022. In particular, Social
Sciences including Arts and Humanties journals accounted for the
strongest growth in the five-years period from 6.4\% (n = 8,361) to 23\%
(n = 51,938), followed by the Life Sciences from 7.6\% (n = 15,003) to
18\% (n = 39,494) , Health Sciences from 5.3\% (n = 18,279) to 16\% (n =
63,089) and Physical Sciences from 4.5\% (n = 22,364) to 12\% (n =
85,428). Growth in Social Sciences category can be largely attributed to
transformative agreements. In 2022, two-thirds of open access articles
(67\%, n = 34,759) were published by the first authors affiliated with
participating institutions (see \ref{fig:subject_panel}B).

@@

\caption{Development of open access in hybrid journals in transformative agreements between 2018 and 2022 by ASJC subject area. The blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established. (A) Proportion of open access articles in hybrid journals per year and subject area (B) Percentage of hybrid open access via agreements per year and subject area Boxplots (C) show individual open access uptake rates by individual hybrid journals and open access funding per publication year and subject area The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available for this study.}\label{fig:subject_panel}
\end{figure}

@@

Between 2018 and 2022, high-income countries almost exclusively
dominated hybrid open access publishing through transformative
agreements. To discern socio-economic differences, these countries were
grouped according to their membership in the Organisation for Economic
Co-operation and Development (OECD). During 2018 and 2022, first authors
affiliated with institutions from OECD member countries published
602,050 open access articles in hybrid journals, representing 81\% of
the investigated open access articles. This disparity between OECD
nations and other countries becomes even more evident when considering
open access through transformative agreements, as 310,712 of 328,957, or
94\% of open access articles were associated with such agreements.

@@

\caption{Development of hybrid open access publishing by country. (A) presents the relative number of articles published in hybrid journals included in transformative agreement by year, distinguishing between BRICS as of 2022, OECD and other countries. (B) Scatterplot contrasts total articles with open access article volume in 2022, by country and its OECD (purple-colored points) or BRICS (green) membership. Trend line obtained from linear regression, grey areas show pointwise symmetric 95\% confidence bands. (C) Hybrid open access market share in 2022 by country. In (A) and (C), the blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established. The remainder shows closed access articles. Country names are represented as ISO two-letter country codes.}\label{fig:country_patch}
\end{figure}

Figure \ref{fig:country_patch}A shows the development of hybrid open
access publishing by country, comparing the OECD area with the BRICS, an
intergovernmental organisation, which comprised Brazil, Russia, India,
China, and South Africa as of 2022. The residual category ``Other''
includes the remaining countries. Full counting was applied to account
for multiple country affiliations (Hottenrott et al., 2021). Open access
in hybrid journals increased in particular because of publications from
authors affiliated with institutions from the OECD area, from 52,154 in
2018 to 202,787 in 2022. This growth was largely driven by
transformative agreements. The share of open access by agreement rose
from 5.5\% in 2018 (n = 2,859) to 68\% (n = 137,815) in 2022. In
contrast, BRICS recorded a low uptake, moderately growing from 1.6\% in
2018 to 3.7\% in 2022.

@@

To illustrate the situation in 2022, Figure \ref{fig:country_patch}B
compares the total publication output with the number of open access
articles. With 391,530 articles, China was the most productive country,
followed by the United States (n = 268,965) and India (n = 87,428). In
contrast, Western and Northern European countries published a
considerably high number of open access articles. Particularly,
Scandinavian countries, the Netherlands and Austria recorded
above-average open access shares, as indicated by the linear trend line.
As shown in Figure \ref{fig:country_patch}C, transformative agreements
contributed to these market positions. Interestingly, the United States
had a notable open access market share of 15\%, although transformative
agreements contributed to a lesser extent. Similarly, China's open
access market share of 7.2\% in 2022 was comparable to that of the
Netherlands, which was 7.1\%.

@@

A key finding of this analysis is that transformative agreements
maintain market concentration. Specifically, the three largest
commercial publishers Elsevier, Springer Nature, and Wiley dominate,
particularly with regard to open access provided through transformative
agreements. Together, the three publishers accounted for three-fourths
of open access articles through transformative agreements, while
recording less than half of the total publication volume published
between 2018 and 2022. This observation aligns with previous research on
shifts in the publishing market following the introduction of funding
opportunities for hybrid open access (Butler et al., 2023; Jahn \&
Tullney, 2016; Shu \& Larivière, 2023). Additionally, the results
confirm previously observed variations by publisher, with Elsevier
exhibiting a different development than Springer Nature and Wiley
(Butler et al., 2023; Jahn et al., 2022). However, it must be noted that
the focus of transformative agreements on publishers with large hybrid
journal portfolios is intentional (Campbell et al., 2022). Because of
transformative agreements with a few large publishers, national
consortia were able to substantially increase their country's annual
open access article volume (Bosman, Jonge, et al., 2021; Huang et al.,
2020; Pinhasi et al., 2021; Taubert et al., 2023; Wenaas, 2022).

@@

Surprisingly, the total publication output of high-income countries
belonging to the OECD in hybrid journals declined substantially after
peaking in 2020, whereas that of BRICS countries doubled from 2018 to
2022. Because the BRICS expansion can be exclusively attributed to
closed access articles, this development has the potential to hinder the
transition of academic journal publishing to full open access through
transformative agreements and demands discussion. In China, the country
with the highest volume of articles in hybrid journals, only limited
research funding to pay for open access is available, with expenditures
for open access publishing surpassing subscription costs (Shu \&
Larivière, 2023). Furthermore, the focus of Chinese authors on
established journals may contribute to this trend (Zhang et al., 2022).
Although China supports Plan S (Schiermeier, 2018), this is not
reflected in the data in terms of open access uptake in hybrid journals.
Market observers also do not expect a broad implementation of
transformative agreements in China in near future (Owens, 2024). As
highlighted by Koley \& Lala (2023), India faces similar challenges in
terms of the availablility of resources to pay for open access. At the
same time, access to research literature is a pressing issue, which is
addressed by the ``Indian one nation, one subscription'' policy
proposal. However, this policy focuses on centrally negotiated
subscriptions and does not entail open access.

@@

This large-scale study provides the first empirical evidence of the
influence of transformative agreements on the transition of hybrid
journals to full open access. However, several limitations need to be
acknowledged. From a data perspective, estimations of open access
through transformative agreements were established by linking author
affiliations with publicly available agreement data from cOAlition S,
and not through invoicing data, which is usually not shared. Because
data on corresponding authors, who are typically responsible for
facilitating open access publication, were not fully supported by
OpenAlex at the time of this study, first author affiliations were used
instead. This study is also unable to account for the various types of
transformative agreements due to a lack of data, especially regarding
article types and caps limiting the number of open access articles
covered. Furthermore, assessing the quality of the OpenAlex and Crossref
data used, particularly in terms of affiliations, corresponding
authorship, and article types, by combining them with established
bibliometric databases such as Scopus and Web of Science was beyond the
scope of this analysis (Chinchilla-Rodríguez et al., 2024; Visser et
al., 2021). It must therefore be emphasised that, unlike related
evaluations of national consortia based on invoice data (Brayman et al.,
2024), this global overview can only provide an estimate of the support
for open access through transformative agreements. Despite these
limitations, the methodology was designed to underestimate, rather than
overestimate, the adoption of open access through transformative
agreements. To promote transparency, the data used in this study, along
with the code used for this analysis, are openly available.

@@

\bibitem[\citeproctext]{ref-Schmal_2024}
Schmal, W. B. (2024). How transformative are transformative agreements?
Evidence from germany across disciplines. \emph{Scientometrics}.
\url{https://doi.org/10.1007/s11192-024-04955-y}

>>>>

The controversies surrounding hybrid open access and transformative
agreements have led to varying policy conclusions. For instance, the
British Joint Information Systems Committee (JISC) evaluated its open
access strategy, which included transformative agreements (Brayman et
al., 2024). The evaluation revealed that while these agreements had a
significant impact on the growth of open access in the country, they had
limited effects in facilitating a global shift towards full open access.
As a result, the report advised that the British open access strategy
should be reassessed. After making similar observation, Norwegian and
Swedish universities and their consortia also argued for policy changes
(Holden et al., 2023; Widding, 2024). Furthermore, cOAlition S conclude
its financial support of transformative agreements at the end of 2024,
but continue to view open access resulting from such agreements as
compliant. \footnote{\url{https://www.coalition-s.org/coalition-s-confirms-the-end-of-its-financial-support-for-open-access-publishing-under-transformative-arrangements-after-2024/}}.
cOAlition S also removed the majority of hybrid journals from its
Transformative Journal program in 2023 due to publishers' failure to
meet self-defined open access growth targets (Brainard, 2023). By
contrast, the German DEAL consortium announced a five-year
transformative agreement with Elsevier starting in 2024 and also renewed
its contracts with Springer Nature and Wiley until the end of 2028.
Similarly, the Colombia Consortium signed the first transformative
agreements in Latin America (Muñoz-Vélez et al., 2024).

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the uptake of open access in hybrid
journals, and the extent to which this can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Brayman et al., 2024; Haucap et al., 2021; Huang et al.,
2020; Pölönen et al., 2020; Taubert et al., 2023; Waltman \& Lamers,
2022; Wenaas, 2022) or publisher portfolios (Bakker et al., 2024; Fraser
et al., 2023; Jahn et al., 2022; Momeni et al., 2023; Pieper \&
Broschinski, 2018; Schmal, 2024), while large-scale studies relied on
self-reported agreement data (Moskovkin et al., 2022), or used APC
pricing lists (Shu \& Larivière, 2023). In particular, data availability
is a limiting factor when studying the impact of transformative
agreements (Bakker et al., 2024), because bibliometric databases, even
though many allow the retrieval of open access articles in hybrid
journals, do not directly attribute them to specific transformative
agreements. Likewise, article-level open access invoicing and cost data,
which would make it possible to establish a direct link between
transformative agreements and open access publications (Jahn et al.,
2022; Kramer, 2024).

@@

This study combines data from multiple publicly available sources, as
shown in Figure \ref{fig:data_workflow}. Initially, transformative
agreement data retrieved from cOAlition S Journal Checker
Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
provided information about journal portfolios and participating
institutions. After identifying hybrid journals by excluding fully open
access journals, Crossref (Hendricks et al., 2020) served as the primary
data source for article-level metadata including Creative Commons (CC)
license information to indicate open access availability on publisher
websites. Because of a lack of comprehensive publicly available
invoicing data, open access articles published through transformative
agreements were determined by linking first author affiliations from
OpenAlex (Priem et al., 2022) to eligible institutions according to the
transformative agreement data. In the following, the steps are described
in detail.

@@

Data gathering started with obtaining journals included in
transformative agreements from the publicly available Transformative
Agreement Data dump\footnote{\url{https://journalcheckertool.org/transformative-agreements/}}
used by the cOAlition S Journal Checker Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}},
a voluntary effort based on publicly disclosed contracts in the ESAC
Transformative Agreement Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}.
The dump consists of multiple online Google spreadsheets where each data
file represents one agreement listed in the ESAC Transformative
Agreement Registry. From the retrieved spreadsheet files, journals and
institutions involved per agreement were obtained.

It should be noted that although many library consortia see the need to
register their agreements through the ESAC registry, some fail to
disclose all details, including the full-text of contracts (Bakker et
al., 2024; Kramer, 2024). Furthermore, the extent to which ESAC is
comprehensive remains uncertain, potentially limiting the coverage of
transformative agreements in the Journal Checker Tool. Another
limitation of using the Journal Checker Tool and its underlying publicly
available data dump to study the development of transformative
agreements over time is that expired transformative agreements are
constantly removed. To address this, four different snapshots were
safeguarded and combined for this study: self-archived versions from
July 2021, July 2022, and May 2023, as well as the most current dump
downloaded on 11 December 2023. This ensured that transformative
agreements, which ended from 2021 onwards, were included, representing
the majority of transformative agreements. Overall, the four combined
Transformative Agreement Data dumps used in this study contained 729 of
869 agreements listed in the ESAC Transformative Agreement Registry by
December 2023.

@@

Affiliation metadata about corresponding authors are crucial for the
planning and evaluation of transformative agreements because they are
considered responsible for arranging open access publication (Borrego et
al., 2021; Geschuhn \& Stone, 2017; Schimmer et al., 2015). For this
study, country and institutional affiliations were retrieved from
OpenAlex. Because the corresponding authorship field was not fully
supported by OpenAlex at the time of analysis, and respective
affiliation data were only available for 54\% of the investigated
articles, this study focused on first authors and their affiliations
instead; approximately 90\% of the articles examined had first author
affiliation metadata in OpenAlex, which is a much larger proportion than
previously reported for the October 2022 snapshot (Zhang et al., 2024).
First authors typically contribute the most to a paper and are often
considered as lead authors (Chinchilla-Rodríguez et al., 2024; Larivière
et al., 2016). In case of lacking data about corresponding authors in
bibliometric databases, related studies also utilised first authors as a
proxy to examine open access payments and transformative agreements
(Haucap et al., 2021; Shu \& Larivière, 2023; Zhang et al., 2022).

To estimate the impact of transformative agreements on hybrid open
access, participating institutions from the Transformative Agreement
Data dump, which the cOAlition S crowd-sourced from agreement documents
and consortia, were matched with the first author affiliations recorded
by OpenAlex using the ROR-ID. Matching also considered the duration of
agreements according to the ESAC registry. In total, 502 agreements were
active between 2018 and 2022. Upon inspection, Transformative Agreement
Data did not comprehensively cover associated institutions, such as
university hospitals or institutes of large research organisations such
as the Max Planck Society. To improve the matching, Transformative
Agreement Data were automatically enriched with ROR-IDs from associated
organisations according to OpenAlex's institution entity data.

@@

Between 2018 and 2022, 11,189 out of 12,857 hybrid journals in
transformative agreements published at least one open access article
under a Creative Commons license. These eligible 11,189 hybrid journals
constituted the foundation for the subsequent analyses. Collectively,
they provided open access to 742,369 out of 8,146,958 articles during
the investigated period, representing a five-year open access proportion
of 9.1\%. Authors who could make use of transformative agreements at the
time of publication contributed 328,957 (44\%) open access articles.
Overall, this investigation was able to establish a link between open
access articles and eligible institutions for 394 out of 502 (78\%)
active transformative agreements between 2018 and 2022 through the
linking of first author affiliations with agreement data.

@@

Figure \ref{fig:results_overview}C depicts the substantial variations
among the hybrid journals included in transformative agreements in terms
of open access uptake. Although the median generally follows the trend
shown in Figure \ref{fig:results_overview}A, the farther stretch of
upper quartiles and whiskers over the years illustrates that an
increasing number of journals published an above-average proportion of
open access articles. In 2022, 25\% of hybrid journals (n = 2,576) had
an open access uptake of 29\%, and 6.6\% of journals (n = 744) provided
the majority of their articles under a CC license in the same year. On
average, these journals were smaller (M = 75, SD = 186) than those with
an open access share below 50\% (M = 164, SD = 347).

@@

However, there are notable differences between the three large
publishers. Although Elsevier published the largest volume of articles
(n = 2,770,826; 34\%), it recorded a comparably low number of open
access articles, including those that are associated with transformative
agreements. In contrast, Springer Nature and Wiley provided open access
to a larger proportion of their articles (13\% of Springer Nature
articles and 15\% of Wiley articles were open access), leading to higher
open access market shares (23\% Springer Nature resp. 23\% Wiley). This
difference between Elsevier on the one hand and Springer Nature and
Wiley on the other can be attributed to transformative agreements, as
the latter made the majority of their open access articles available
through such deals (57\% Springer Nature resp. 55\% Wiley).

@@

The varying degrees of adoption of open access across the three major
publishers can be attributed to distinct approaches to transformative
agreements. Springer Nature, for example, started offering open access
agreements for its hybrid journal portfolio to selected consortia such
as the Max Planck Society, the Swedish Bibsam consortium and the Finnish
FinELib consortium in 2015 under the name Springer Compact\footnote{\url{https://web.archive.org/web/20180414062853id_/http://www.liber2015.org.uk/wp-content/uploads/2015/03/Springer-Compact.pdf}}.
Prior to this, Springer had some pilot agreements with a small number of
institutions, including the University of Göttingen (Schmidt \& Shearer,
2012). However, the Springer Compact agreements were not included in the
data as they concluded before the start of the transformative agreement
data collection in June 2021. Nonetheless, the results demonstrate the
importance of agreements for Springer Nature's hybrid open access
business over the five-years period (Figure
\ref{fig:publisher_figure}B). In 2022, 66\% (n = 34,725) of open access
in Springer Nature hybrid journals was enabled through transformative
agreements. In the same year, 70\% (n = 37,316) of Wiley's open access
articles could be linked to transformative agreements. By contrast,
Elsevier published a comparatively lower proportion of its open access
articles through transformative agreements in 2022 (n = 32,627; 54\%).

@@

Figure \ref{fig:subject_panel} presents the relative growth of hybrid
open access by subject area between 2018 and 2022. In particular, Social
Sciences including Arts and Humanities journals accounted for the
strongest growth in the five-years period from 6.4\% (n = 8,361) to 23\%
(n = 51,938), followed by the Life Sciences from 7.6\% (n = 15,003) to
18\% (n = 39,494) , Health Sciences from 5.3\% (n = 18,279) to 16\% (n =
63,089) and Physical Sciences from 4.5\% (n = 22,364) to 12\% (n =
85,428). Growth in Social Sciences category can be largely attributed to
transformative agreements. In 2022, two-thirds of open access articles
(67\%, n = 34,759) were published by the first authors affiliated with
participating institutions (see \ref{fig:subject_panel}B).

@@

\caption{Development of open access in hybrid journals in transformative agreements between 2018 and 2022 by ASJC subject area. The blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established. (A) Proportion of open access articles in hybrid journals per year and subject area. (B) Percentage of hybrid open access via agreements per year and subject area. (C) Boxplots show individual open access uptake rates by individual hybrid journals and open access funding per publication year and subject area. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available for this study.}\label{fig:subject_panel}
\end{figure}

@@

Between 2018 and 2022, high-income countries almost exclusively
dominated hybrid open access publishing through transformative
agreements. To discern socio-economic differences, these countries were
grouped according to their membership in the Organisation for Economic
Co-operation and Development (OECD). Overall, first authors affiliated
with institutions from OECD member countries published 602,050 open
access articles in hybrid journals, representing 81\% of the
investigated open access articles. This disparity between OECD nations
and other countries becomes even more evident when considering open
access through transformative agreements, as 310,712 of 328,957, or 94\%
of open access articles were associated with such agreements.

@@

\caption{Development of hybrid open access publishing by country. (A) presents the relative number of articles published in hybrid journals included in transformative agreement by year, distinguishing between BRICS as of 2022, OECD and other countries. (B) Scatterplot contrasts total articles with open access article volume in 2022, by country and its OECD (purple-colored points) or BRICS (green) membership. Trend line obtained from linear regression, grey area show pointwise symmetric 95\% confidence bands. (C) Hybrid open access market share in 2022 by country. In (A) and (C), the blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established. The remainder shows closed access articles. Country names are represented as ISO two-letter country codes.}\label{fig:country_patch}
\end{figure}

Figure \ref{fig:country_patch}A shows the development of hybrid open
access publishing by country, comparing the OECD area with the BRICS, an
intergovernmental organisation, which comprised Brazil, Russia, India,
China, and South Africa as of 2022. The residual category ``Other''
includes the remaining countries. Full counting was applied to account
for multiple country affiliations (Hottenrott et al., 2021). Open access
in hybrid journals increased in particular because of publications from
first authors affiliated with institutions from the OECD area, from
52,154 in 2018 to 202,787 in 2022. This growth was largely driven by
transformative agreements. Their share rose from 5.5\% in 2018 (n =
2,859) to 68\% (n = 137,815) in 2022. In contrast, BRICS recorded a low
uptake, moderately growing from 1.6\% in 2018 to 3.7\% in 2022.

@@

To illustrate the situation in 2022, Figure \ref{fig:country_patch}B
compares the total publication output with the number of open access
articles. With 391,530 articles, China was the most productive country,
followed by the United States (n = 268,965) and India (n = 87,428). In
contrast, Western and Northern European countries published a
considerably high number of open access articles. Particularly, Nordic
countries, the Netherlands and Austria recorded above-average open
access shares, as indicated by the linear trend line. As shown in Figure
\ref{fig:country_patch}C, transformative agreements contributed to these
market positions. Interestingly, the United States had a notable open
access market share of 15\%, although transformative agreements
contributed to a lesser extent. Similarly, China's open access market
share of 7.2\% in 2022 was comparable to that of the Netherlands, which
was 7.1\%.

@@

A key finding of this analysis is that transformative agreements
maintain market concentration. Specifically, the three largest
commercial publishers Elsevier, Springer Nature, and Wiley dominate,
particularly with regard to open access provided through transformative
agreements. Together, the three publishers accounted for three-fourths
of open access articles through transformative agreements, while
recording about half of active hybrid journals included in
transformative agreements between 2018 and 2022. This observation aligns
with previous research on shifts in the publishing market following the
introduction of funding opportunities for hybrid open access (Butler et
al., 2023; Jahn \& Tullney, 2016; Shu \& Larivière, 2023). Additionally,
the results confirm previously observed variations by publisher, with
Elsevier exhibiting a different development than Springer Nature and
Wiley (Butler et al., 2023; Jahn et al., 2022). However, it must be
noted that the focus of transformative agreements on publishers with
large hybrid journal portfolios is intentional (Campbell et al., 2022).
Because of transformative agreements with a few large publishers,
national consortia were able to substantially increase their country's
annual open access article volume (Bosman, Jonge, et al., 2021; Brayman
et al., 2024; Huang et al., 2020; Pinhasi et al., 2021; Taubert et al.,
2023; Wenaas, 2022; Widding, 2024).

@@

Surprisingly, the total publication output of high-income countries
belonging to the OECD in hybrid journals declined substantially after
peaking in 2020, whereas that of BRICS countries continuously grew from
2018 to 2022. Because the BRICS expansion can be exclusively attributed
to closed access articles, this development has the potential to hinder
the transition of academic journal publishing to full open access
through transformative agreements and demands discussion. In China, the
country with the highest volume of articles in hybrid journals, only
limited research funding to pay for open access is available, with
expenditures for open access publishing surpassing subscription costs at
some universities (Shu \& Larivière, 2023). Furthermore, the focus of
Chinese authors on established journals may contribute to this trend
(Zhang et al., 2022). Although China supports Plan S (Schiermeier,
2018), this is not reflected in the data in terms of open access uptake
in hybrid journals. Market observers also do not expect a broad
implementation of transformative agreements in China in near future
(Owens, 2024). As highlighted by Koley \& Lala (2023), India faces
similar challenges in terms of the availability of resources to pay for
open access. At the same time, access to research literature is a
pressing issue, which is addressed by the ``Indian one nation, one
subscription'' policy proposal. However, this policy focuses on
centrally negotiated subscriptions and does not entail open access.

@@

This large-scale study provides the first empirical evidence of the
influence of transformative agreements on the transition of hybrid
journals to full open access. However, several limitations need to be
acknowledged. From a data perspective, estimations of open access
through transformative agreements were established by linking author
affiliations with publicly available agreement data from cOAlition S,
and not through invoicing data, which is usually not shared. Because
data on corresponding authors, who are typically responsible for
facilitating open access publication, were not fully supported by
OpenAlex at the time of this study, first author affiliations were used
instead. This study is also unable to account for the various types of
transformative agreements due to a lack of data, especially regarding
article types and caps that limit the number of open access articles
covered. Furthermore, assessing the quality of the OpenAlex and Crossref
data used, particularly in terms of affiliations, corresponding
authorship, and article types, by combining them with established
bibliometric databases such as Scopus and Web of Science was beyond the
scope of this analysis (Chinchilla-Rodríguez et al., 2024; Visser et
al., 2021). It must therefore be emphasised that, unlike evaluations
from national consortia that could make use of invoices (Brayman et al.,
2024), this global overview can only provide an estimate of the support
for open access through transformative agreements. Despite these
limitations, the methodology was designed to underestimate, rather than
overestimate, the adoption of open access through transformative
agreements. To promote transparency, the data used in this study, along
with the code used for this analysis, are openly available.

@@

\bibitem[\citeproctext]{ref-Schmal_2024}
Schmal, W. B. (2024). How transformative are transformative agreements?
Evidence from {Germany} across disciplines. \emph{Scientometrics},
\emph{129}(3), 1863--1889.
\url{https://doi.org/10.1007/s11192-024-04955-y}
